---
title: "DLBCL mutations and chromosomal alterations"
author: "Matthew Bendall"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: yes
    toc_depth: 3
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
  html_notebook:
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: paper
    toc: yes
    toc_depth: 3
    code_folding: hide
    toc_float:
      collapsed: no
      smooth_scroll: yes
params:
  dedup: yes
  mut_threshold: 4
---

```{css, echo=FALSE}
.scroll-150 {
  max-height: 150px;
  overflow-y: auto;
  background-color: inherit;
}
```

# Mutations and Chromosomal Alterations {.tabset}

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
root.dir <- rprojroot::find_rstudio_root_file()

library(tidyverse)
```

### Load clusters

```{r}
load("r_outputs/05o-DLBCL_pca_ccp_clusters_metadata.Rdata")

dlbcl.tbl <- DLBCL_metadata %>% 
  rownames_to_column() %>%
  as_tibble %>%
  mutate(
    subtype.herv = factor(clust.retro.k7, levels = str_c('HC', 1:7))
  )
```

##### sanity check

First check that the order of the sample names (`dlbcl.tbl$rowname`) is consistent
among `dlbcl.tbl`, `clust.df`, `cDat`, and `ccp.obj`.

Then check if the cluster assignments (`dlbcl.tbl$subtype.herv`) are the same
in `dlbcl.tbl`, `clust.df`, and `ccp.obj`.

```{r}
# sanity check
stopifnot(
    all(dlbcl.tbl$rowname == rownames(clust.df)),
    all(dlbcl.tbl$rowname == colnames(cDat)),
    all(dlbcl.tbl$rowname == names(ccp.obj[[7]]$consensusClass))
)

stopifnot(
    all(str_c('H', dlbcl.tbl$subtype.herv) == clust.df$clust.retro.k7),
    all(as.integer(dlbcl.tbl$subtype.herv) == ccp.obj[[7]]$consensusClass)
)
```

#### Verify tree is same as in manuscript

```{r}
dlbcl.tree <- ccp.obj[[7]]$consensusTree
plot(as.dendrogram(dlbcl.tree))
```

## MAF data (mutations)

```{r}
maf.nci.fn <- 'f0325846-10a5-46b9-984d-c93a473bb6a4/MAF_NCICCR-DLBCL_phs001444.txt'
maf.tcga.fn <- '7156a0a2-4df1-446f-a8cd-31df5b121e74/MAF_TCGA-DLBC_phs000178.txt'

maf.tbl.orig <- rbind(
  (
    read.table(
      file.path('metadata/Schmitz_GDC', maf.nci.fn), header=T, sep='\t'
    ) %>% mutate(dataset=factor('NCICCR', levels=c('NCICCR','TCGA')))
  ),
  (
    read.table(
      file.path('metadata/Schmitz_GDC', maf.tcga.fn), header=T, sep='\t'
    ) %>% mutate(dataset=factor('TCGA', levels=c('NCICCR','TCGA')))
  )
) %>% as_tibble() %>%
  mutate(MUTATION_TYPE=factor(MUTATION_TYPE, levels=c("TRUNC", "INFRAME", "MISSENSE"))) %>%
  dplyr::relocate(SUBJECT_NAME, MUTATION_ID, GENE.SYMBOL, AA.CHANGE, MUTATION_TYPE, DETECTION_TYPE) %>%
  select(-c(dbGaP_ACCESSION))
```

```{r include=FALSE}
rm(maf.nci.fn, maf.tcga.fn)
```

### Original MAF table

```{r echo=FALSE, rows.print=20}
maf.tbl.orig
```

```{r eval=FALSE, include=FALSE}
# Subject name, mutation ID, detection type together are sufficent.
(maf.tbl.orig %>% nrow()) == (maf.tbl.orig %>% select(SUBJECT_NAME, MUTATION_ID, DETECTION_TYPE) %>% distinct() %>% nrow())
```

#### Dedup

Remove duplicate subject+mutation combinations that differ only by detection
type. Depends on `params$dedup`.

```{r}
if(params$dedup) {
  maf.tbl.dup <- maf.tbl.orig %>% 
    filter(duplicated(maf.tbl.orig %>% select(SUBJECT_NAME, MUTATION_ID)))
  maf.tbl <- maf.tbl.orig %>% select(-DETECTION_TYPE) %>% distinct()
} else {
  maf.tbl <- maf.tbl.orig
}
```

### MAF table

```{r echo=FALSE, rows.print=20}
maf.tbl
```

```{r eval=FALSE, include=FALSE}
dup.i <- 246
maf.tbl.orig %>%
  filter(
    SUBJECT_NAME == maf.tbl.dup[dup.i,]$SUBJECT_NAME & MUTATION_ID == maf.tbl.dup[dup.i,]$MUTATION_ID
  )

maf.tbl %>%
  filter(
    SUBJECT_NAME == maf.tbl.dup[dup.i,]$SUBJECT_NAME & MUTATION_ID == maf.tbl.dup[dup.i,]$MUTATION_ID
  )

```

#### Number of mutations per sample

```{r rows.print=10}
count.bysample <- maf.tbl %>% 
  group_by(SUBJECT_NAME) %>% 
  summarise(
    nMUT = n_distinct(MUTATION_ID),
    uGENE = n_distinct(GENE.SYMBOL)
  ) %>% 
  arrange(desc(nMUT))

count.bysample
```

### Collapse by subject+gene

```{r}
# Keeps all rows of maf.tbl and adds counts per sample and gene, combines
# AA.CHANGE and MUTATION_TYPE into a comma-separated list. Mutations for
# genes with multiple mutations are sorted by mutation type and position.
annot.maf.tbl <- (
  maf.tbl %>%
    # mutate(
    #   sort.AA.CHANGE = as.integer(str_extract(AA.CHANGE, "[0-9]+"))
    # ) %>%
    # mutate(
    #   sort.AA.CHANGE = ifelse(is.na(sort.AA.CHANGE), 0, sort.AA.CHANGE)
    # ) %>%
    group_by(SUBJECT_NAME) %>%
    mutate(nmut_sample=n()) %>%
    group_by(SUBJECT_NAME, GENE.SYMBOL) %>%
    mutate(nmut_gene = n()) %>%
    arrange(MUTATION_TYPE, HG.19.POSITION, .by_group = TRUE) %>%
    mutate(
      AA.CHANGE.slst = str_c(unlist(AA.CHANGE), collapse=','),
      MUTATION_TYPE.slst = str_c(unlist(MUTATION_TYPE), collapse=',')
    )
)

# One row per subject+
collapse.gene.tbl <- (
  annot.maf.tbl %>%
    slice_head(n=1) %>%
    arrange(desc(nmut_sample), desc(nmut_gene))
)
```

```{r eval=FALSE, include=FALSE}
annot.maf.tbl %>% filter(SUBJECT_NAME=='DLBCL10478' & GENE.SYMBOL=='ARHGEF1')
```

```{r eval=FALSE, include=FALSE}
collapse.gene.tbl %>% filter(SUBJECT_NAME=='DLBCL10478' & GENE.SYMBOL=='ARHGEF1')
```

```{r rows.print=5}
collapse.gene.tbl %>% 
  select(SUBJECT_NAME, GENE.SYMBOL, AA.CHANGE, AA.CHANGE.slst, MUTATION_TYPE, MUTATION_TYPE.slst)
```

```{r eval=FALSE, include=FALSE}
# check that collapse.gene.tbl agrees with count.bysample
all(sapply(sample(1:nrow(count.bysample), 15), function(rn) {
  tmp <- collapse.gene.tbl %>% filter(SUBJECT_NAME == count.bysample$SUBJECT_NAME[rn]) 
  (nrow(tmp) == count.bysample$uGENE[rn]) & (sum(tmp$nmut_gene) == count.bysample$nMUT[rn])
}))
```

### Extract mutations to table

```{r}
all.genes <- collapse.gene.tbl %>% group_by(GENE.SYMBOL) %>% tally(sort=T)
all.genes
```

```{r}
snames <- setNames(nm=dlbcl.tbl$case)
gnames <- (all.genes %>% filter(n > params$mut_threshold))$GENE.SYMBOL %>% setNames(nm=.)
```


```{r}
snames[!(snames %in% collapse.gene.tbl$SUBJECT_NAME)]
```

```{r}
mut.detail.df <- {
  df.lst <- lapply(snames, function(s) {
    stbl <- collapse.gene.tbl %>% ungroup() %>% filter(SUBJECT_NAME == s)
    left_join(
      tibble(GENE.SYMBOL=gnames),
      stbl %>% select(GENE.SYMBOL, AA.CHANGE, nmut_gene, AA.CHANGE.slst),
      by=join_by(GENE.SYMBOL)
    ) %>%
      mutate(val=AA.CHANGE.slst) %>%
      pull(val, name=GENE.SYMBOL) %>%
      {if(nrow(stbl)>0) replace_na(., ".") else .}
  })
  df.lst %>% bind_rows() %>% data.frame(row.names=names(df.lst))
}

mut.detail.df
```


```{r}
mut.top.df <- {
  df.lst <- lapply(snames, function(s) {
    stbl <- collapse.gene.tbl %>% ungroup() %>% filter(SUBJECT_NAME == s)
    left_join(
      tibble(GENE.SYMBOL=gnames),
      stbl %>% select(GENE.SYMBOL, AA.CHANGE, nmut_gene, AA.CHANGE.slst),
      by=join_by(GENE.SYMBOL)
    ) %>%
      mutate(val=AA.CHANGE) %>%
      pull(val, name=GENE.SYMBOL) %>%
      {if(nrow(stbl)>0) replace_na(., ".") else .}
  })
  df.lst %>% bind_rows() %>% data.frame(row.names=names(df.lst))
}

mut.top.df
```


```{r}
mut.simple.df <- {
  df.lst <- lapply(snames, function(s) {
    stbl <- collapse.gene.tbl %>% ungroup() %>% filter(SUBJECT_NAME == s)
    left_join(
      tibble(GENE.SYMBOL=gnames),
      stbl %>% select(GENE.SYMBOL, AA.CHANGE, nmut_gene, AA.CHANGE.slst) %>% mutate(val='x'),
      by=join_by(GENE.SYMBOL)
    ) %>%
      pull(val, name=GENE.SYMBOL) %>%
      {if(nrow(stbl)>0) replace_na(., ".") else .}
  })
  df.lst %>% bind_rows() %>% data.frame(row.names=names(df.lst))
}

mut.simple.df
```

### Write RDS

```{r}
saveRDS(mut.detail.df, file = 'r_outputs/08-DLBCL_mutations/mut.detail.df.Rds')
saveRDS(mut.top.df, file = 'r_outputs/08-DLBCL_mutations/mut.top.df.Rds')
saveRDS(mut.simple.df, file = 'r_outputs/08-DLBCL_mutations/mut.simple.df.Rds')
```

## Translocation data

```{r}
fus.nci.fn <- '3e803605-41bf-45b3-9b55-59c4c2610c60/Fusion_BCL6_BCL2_MYC-NCICCR-DLBCL_phs001444.xlsx'
fus.tcga.fn <- 'b5deda7e-9971-4426-9bc0-bad025ad6d38/Fusion_BCL6_BCL2_MYC-TCGA-DLBC_phs000178.txt.xlsx'

fus.tbl.orig <- rbind(
  readxl::read_xlsx(file.path('metadata/Schmitz_GDC', fus.nci.fn)),
  readxl::read_xlsx(file.path('metadata/Schmitz_GDC', fus.tcga.fn))
) %>%
  as_tibble() %>%
  mutate(
      fusion.BCL6 = replace_na(`BCL6 locus fusion`, "."),
      fusion.BCL2 = replace_na(`BCL2  locus fusion`, "."),
      fusion.MYC = replace_na(`MYC locus fusion`, ".")
  ) %>%
  select(SUBJECT_NAME=`dbGAP subject ID`, fusion.BCL6, fusion.BCL2, fusion.MYC)
```


```{r include=FALSE}
rm(fus.nci.fn, fus.tcga.fn)
```

### Original MAF table

```{r echo=FALSE, rows.print=20}
fus.tbl.orig
```
### Join

```{r}
fus.df <- left_join(
  dlbcl.tbl %>% select(case),
  fus.tbl.orig,
  by = join_by(case == SUBJECT_NAME)
) %>% column_to_rownames("case")

fus.df
```

### Write RDS

```{r}
saveRDS(fus.df, file = 'r_outputs/08-DLBCL_mutations/fus.df.Rds')
```
