---
title: "DLBCL mutations - example analysis"
author: "Matthew Bendall"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    github_document:
        toc: true
        toc_depth: 3
    html_notebook:
        df_print: paged
        highlight: tango
        number_sections: true
        theme: paper
        toc: yes
        toc_depth: 3
        code_folding: hide
        toc_float:
            collapsed: no
            smooth_scroll: yes
---

```{css, echo=FALSE}
.scroll-150 {
  max-height: 150px;
  overflow-y: auto;
  background-color: inherit;
}
```

# Statistical Testing (example) {.tabset}

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
root.dir <- rprojroot::find_rstudio_root_file()

library(tidyverse)
library(magrittr)
```


## Load and wrangle data

```{r}
load("r_outputs/05o-DLBCL_pca_ccp_clusters_metadata.Rdata")

dlbcl.tbl <- DLBCL_metadata %>% 
  rownames_to_column() %>%
  as_tibble %>%
  mutate(
    subtype.herv = factor(clust.retro.k7, levels = str_c('HC', 1:7))
  )

mut.simple.df <- readRDS('r_outputs/08-DLBCL_mutations/mut.simple.df.Rds')
fus.df <- readRDS('r_outputs/08-DLBCL_mutations/fus.df.Rds')

# factorize
mut.simple.df %<>% mutate(across(matches("."), ~ factor(.x, levels=c('.', 'x'))))
fus.df %<>% mutate(across(matches("."), ~ factor(.x, levels=c('.', 'x'))))

mutfus.tbl <- (
    dlbcl.tbl %>%
      select(case, subtype.herv) %>%
      left_join(
          mut.simple.df %>% rownames_to_column('case') %>% as_tibble(),
          by="case"
      ) %>%
      left_join(
          fus.df %>% rownames_to_column('case') %>% as_tibble(),
          by="case"
      )
)
```

```{r rows.print=20}
mutfus.tbl
```
## Chi square

###### Is there a significant difference in the distribution of a mutation across the HERV subtypes?

(Could wrap this in an lapply)

```{r}
chisq.myfunc <- function(g) {
      
    cat(paste0("## Gene: ", g, "\n"))
    obs.tbl <- as.matrix(table(mutfus.tbl$subtype.herv, mutfus.tbl[[g]]))
    
    cat("\n## Observed")
    print(obs.tbl)
    
    res <- chisq.test(obs.tbl)
    print(res)
    
    inttext <- paste0(
      "\n## Interpretation:\n",
      "## There is ", 
      ifelse(res$p.value<0.05, "", "not "),
      "a signficant difference in the distribution of ",
      g,
      " mutations\n## across HERV subtypes.\n"
    )
    
    cat(inttext)
    
    # cat("\n## Observed from test")
    # res$observed
    
    cat("\n## Expected")
    print(res$expected)
    
    cat("\n## Residuals")
    print(res$residuals)
    data.frame(res$residuals) %>% 
      mutate(Var2 = ifelse(Var2=='x', 'VAR', 'WT')) %>% 
      pivot_wider(names_from = Var2, values_from=Freq) %>%
      mutate(msg = ifelse(WT<0 & VAR>0, "more mutations", "fewer mutations")) %>%
      mutate(msg = paste0(Var1, ": ", msg, " than expected")) %>%
      pull(msg) %>%
      str_c(collapse='\n##     ') %>% cat('\n##     ', ., '\n', sep = '')

    contrib <- (res$observed - res$expected)^2 / res$expected
    pct.contrib <- (contrib / res$statistic) * 100
    
    cat("\n## Percent contribution")
    print(pct.contrib)
    
    if(res$p.value<0.05) {
      inttext2 <- paste0("\n## Interpretation:")
      pc <- data.frame(pct.contrib) %>% arrange(desc(Freq))
      pc <- {if(max(pc$Freq)>10) pc %>% filter(Freq>10) else pc %>% slice_max(Freq, n=3)}
      msgs <- sapply(1:nrow(pc), function(i) paste0(
        'For ', pc$Var1[i], ", ",
        ifelse(pc$Var2[i]=="x", "a mutation in ", "wildtype "), g, 
        " contributes ", 
        sprintf('%.1f', pc$Freq[i]), 
        "% to the overall signficant test."
      ))
      inttext2 <- paste0(inttext2, str_c(str_c('\n##     ', msgs), collapse=""))
      cat(inttext2)
    }
}
```

```{r}
chisq.myfunc('MYD88')
```

```{r}
chisq.myfunc('fusion.BCL6')
```

## Logistic regression

### Cluster as predictor (variant ~ cluster)

Cluster as the predictor of variant (i.e. variant is the dependent variable, it depends on the cluster assignment).

```{r}
xtabs(~subtype.herv + MYD88, data=mutfus.tbl)
model.logit <- glm(MYD88 ~ subtype.herv, data=mutfus.tbl, family="binomial")
summary(model.logit)
```

### Variant as predictor (cluster ~ variant)

Variant as the predictor of cluster (i.e. cluster is the dependent variable, it 
depends on the variant). Allows testing of multiple variants.

```{r}
xtabs(~MYD88 + subtype.herv, data=mutfus.tbl)
model.logit <- glm(subtype.herv ~ MYD88, data=mutfus.tbl, family="binomial")
summary(model.logit)
```

```{r}
model.logit <- glm(subtype.herv ~ MYD88*CD79B, data=mutfus.tbl, family="binomial")
summary(model.logit)
```

